---
title: "LearningR_oReilly"
output: html_document
---




# Learning R 
# https://www.oreilly.com/library/view/learning-r/9781449357160/


# Classes
 All variables in R have a class, which tells you what kinds of variables they are. 
 For example, most numbers have class numeric (see the next section for the other types), 
 and logical values have class logical. Actually, being picky about it, vectors of numbers
 are numeric and vectors of logical values are logical, since R has no scalar types. 
 The “smallest” data type in R is a vector.
 You can find out what the class of a variable is using class(my_variable):

sequence integers
seq.int(2,3, by=0.01)

# sequence a vector 
```{r}
l = c("one","two","three","four")
for(i in seq_along(l)) 
  print(l[i])

# print the length of vector
print( length(l))
```

# vector length
``length(1:5)``



## Naming vector elements (like Python dict)
```{r named vector elements}
named.vector = c(apple=1,
                 cherry=2,
                 mango=8)

named.vector
```

add to a named vector, need ``names()`` keyword
```{r add to vector}
x = 1:5
names(x) = named.vector

x
```



# Indexing Vectors

```{r}

v = c(1:5)**2
v

v[c(1,3,5)]


v[v>6]
```

using the ``which.min`` and ``which.max``
```{r}
which.min(v)
which.max(v)

```


Vector Recycling & Repetition 
``1:5 +1`` is the same ``1+ 1:5``

add 2 vectors together, concatenates
1:5 + 1:15
return error message when short + long vectors added

equal length vectors are a must, use the  ``rep()`` function

repeat 1:5 3 times
``rep(1:5, 3)``

``rep(1:5, each= 3)``

``rep(1:5, times= 1:5)``

``rep(1:5, length.out = 7)``

faster to code
``rep.int(1:5, 3)`` 

``rep_len(1:5, 15)``

# Matrices & Arrays
```{r matrix}
m = matrix(1:12,
           nrow = 4,
           dimnames = list(
             c('one','two','three','four'), # rows
             c('TNG','DS9','VOY') # columns
           ))
m
```

get the dimensions of matrix
``dim(m)``
number of rows
``nrow(m)``

# Matrix or Array indexing
grab first row
m[1,]
grab first column
m[,1]


```{r}
m2 = matrix(
  seq.int(2,24, 2),
  nrow = 4,
  dimnames = list(
    c('r1','r2','r3','r4'),
    c('Engineering','Holodeck','Bridge')
  )
)
m2
```
Transpose a matrix
``t(m)``

inner matrix multiplication
``m %*% t(m) ``

outer matrix multiplication
``m %o% m``

## Quiz
```{r Quiz}
# get vector that has 0,.25,.50,.75, 1
seq(0,1, by= 0.25)

named.vector

```




# Lists
```{r}
(a.list = list( c(33,78,99,45,35),
                month.abb,
                matrix( c(3,-8,1,-4), nrow = 2),
                asin
                )
 )


```


# row, column, and dimension names
```{r}
rownames(m)
colnames(m)
dimnames(m)
```

# indexing arrays
```{r}
# 1st row, columns 2 and 3
m[1, c("DS9", "VOY")]

m[1,]

```

# combine matrices with cbind and rbind
```{r}

cbind(m, m2)

rbind(m, m2)
```


















