---
title: "Random R Notebook"
output: html_notebook
---

# Bechdel Test 
- 2+ named female chars 
- they talk to each other
- talk about something besides a man

RichardOnData - tuorial 5

# Base R 
search for built-in datasets 
``library(help= "datasets")``

see documentation of library
``help(package = 'dplyr')``

```{r Base R}
# help function opens documentation guide
? mean
```

```{r}
str(iris)
```

## summary function (Python .describe() )
``summary(iris)``

## names functions (Python column names)
``names(iris)``

## package installation
``install.package('dplyr')``

## load library
``library(dplyr)``

## function from a library 
<library> :: <function>
``dplyr::select `

## load built-in library
``data(iris)``

## get working directory
``getwd()``
``setwd()``


# vectors
``c(1,2,3)``

integer sequence
``2:6``

a sequence of numbers with a step
``seq(2,3, by= 0.5)``

repeating a vector
``rep(1:4, times= 3)``

repeat elements in vector by n
``rep(1:2, each= 3)``
or 
``n= 3``
``rep(1:2, each= n)`


# Vector Functions

sort vector
``sort(v)``

reverse vector order
``rev(v)``

counts of values
``table(v)``

see unique values
``unique(v)``


# Select Vector Elements

get the 4th element
``v[4]``

all but the 4th element
``v[-4]``



# for loop
```{r for loop}

for (i in k) {
  y = i**2
  print(y)
}
```

# while loop
```{r while loop}

while(i < 15){
  i = i+2
  print(i)
}
```


# Named function
```{r named function}
square = function(x){
  square = x**x
  print(square)
}
square(3)

```






# Random Normal Distribution 

```{r rnorm()}
# get 20 values, mean= 0, std= 1
rv = rnorm(20, 0, 1)
rv

plot(rv)
```


# Density Function
```{r dnorm()}

dv = dnorm(0, 0, 1)
dv


```

# Probability function
```{r pnorm()}
# probability of getting 1.96
pv = pnorm(1.96, 0, 1)
pv
```

# Quantile function
```{r qnorm()}

qv = qnorm(0.975, 0, 1)
qv
```









# IfElse statement
using the ifelse() to get sepal lengths > 6 and sepal widths >3, 
check for condition, return True or False
```{r}

iris2 = iris %>% 
  mutate(LargeSepal = ifelse(Sepal.Length > 6 & Sepal.Width > 3, T, F))

summary(iris2)
```


# Create a function
This function is temperature conversion from Fahrenheit to Celsius
```{r Function}

fahr =  c(76,78,65,71,55,58,76,81)

fahr.2.cels = function(x){
  y = (x - 32) * (5/9)
  return( round(y))
}

celsius = fahr.2.cels(fahr)
celsius
```



Accessing data frame column values
```{r}
# iris df $ column
iris$Sepal.Length
```

matrix notation for data frames
```{r}
iris[, 1]
```





# Binding columns of data frames
add a vector or data frame to another data frame using cbind()
```{r cbind()}

set.seed(555)
noise = rnorm(nrow(iris), 0,1)

cbin.example = cbind(iris, noise)
str(cbin.example)
```



# Binding rows of data frames
create a new data frame that is similar to iris dataset
```{r rbind()}

set.seed(555)

extraRows = data_frame(
  Sepal.Length = rnorm(50, 5, 1),
  Sepal.Width = runif(50, 2, 6),
  Petal.Length = rnorm(50, 1.5, 0.3),
  Petal.Width = runif(50, 0.1, 0.7),
  Species = factor( sample( c('setosa','versicolor', 'virginica'),
                            size = 50, 
                            replace = T))
)

rbind.example = rbind(iris, extraRows)

str(iris)
str(extraRows)
str(extraRows)
```



# using the ``lm()`` function for Linear Regression

```{r Linear Regression}
# don't break up the string without the escape backslash
Lin.Regr = lm( formula= "Sepal.Length ~ Sepal.Width + \
               Petal.Length + Petal.Width",
               data= iris)
summary(Lin.Regr)
```

# Generalized Linear Models ``glm()``
this is used for Logistic Regression
```{r Log Regression}

Log.Regr = glm( formula = "SepalGroup2 ~ Petal.Length + Petal.Width",
                data = iris, 
                family = 'binomial')

summary(Log.Regr)
```





















































```{r}
x = rnorm(n= 50, mean= 10, sd= 20)
x
plot(x)
t.test(x= x, mu= 0, conf.level = 0.95)

```
95 percent confidence interval:
  4.515878 16.077253
  
  
  
  
  
  
  
  
  
  
```{r}

x = rnorm(n= 50, mean= 10, sd= 20)
t.test(x= x, mu= 4.515878, conf.level = 0.95)

```



# working dates using libridate library
```{r working with dates}
library(lubridate)

date.YMD = "2017-03-28"

lubed.date = lubridate::ymd(date.YMD)
lubed.date

date.MDY = "03-28-2018"
lubed.MDY = mdy(date.MDY)
lubed.MDY
# lubridate::mdy(date.MDY)

date.hms = "03-28-2018 14:23:12"
# lubridate::mdy_hms(date.hms)
lubed.hms = mdy_hms(date.hms)
lubed.hms

# lubridate::date_decimal(2045.18)


# lubridate::dhours(x= 9) # 9 hours into seconds
hours_2_sec = dhours(x= 9)
hours_2_sec

# lubridate::dminutes(x= 6) # 6 minutes into second
min_2_sec = dminutes(x= 9)
min_2_sec

# lubridate::dweeks(x= 2) # weeks into seconds
weeks_2_sec = dweeks(x=2)
weeks_2_sec


# lubridate::dmonths(x= 1) # month in weeks to seconds
monthsIn_wks_sec = dmonths(x=1)
monthsIn_wks_sec

# lubridate::dyears(x=1) # 1 year into seconds
yrInSeconds = dyears(x=1)
yrInSeconds

```

use **mutate** function to convert string format date data from CSV files to datetime format. Example file has column 'date_time'

example_df = read.csv("fake_file.csv")
example_df %>% 
  mutate(date_time = mdy_hms(date_time))
















